---
title: "PebbleCount_Cleaning"
author: "Bethany Blakey"
date: "2025"
output: html_document
---

## Cleaning Raw Pebble Count Data

### Load necessary libraries

```{r}
library(tidyverse)
```

## Raw Data:
### Make a list of the files in the folder with the raw data

Data is in rows and columns, one cell per rock that was measured

```{r}
raw_file_list <- list.files("pc_raw", pattern = "*.csv", full.names=TRUE)
```


### Loop through each file and do the following:
```{r}
for (file in raw_file_list) {
  
  # get the name of the file
  name <- tools::file_path_sans_ext(basename(file))
  
  # read the csv
  data <- read.csv(file, header = FALSE)
  
  # convert to a single row
  size <- as.vector(t(data))
  df_sr <- data.frame(size)
  
  # remove any spaces before or after from entry
  df_sr$size <- trimws(df_sr$size)
  
  # create a frequency table for size
  size <- df_sr$size
  freq_table <- table(size)
  
  # convert to a dataframe
  df <- as.data.frame(freq_table)
  
  # Ensure 'size' is treated as character for processing
  df$size <- as.character(df$size)
  
  for (i in seq_len(nrow(df))) {
    if (df$size[i] == "<300") {
      df$size[i] <- 300
    } else if (df$size[i] == ">300") {
      df$size[i] <- 450
    } else if (df$size[i] == "fine") {
      df$size[i] <- 2
    }
  }
  
  df$size <- suppressWarnings(as.numeric(df$size))
  
  # remove empty values
  df <- df[!is.na(df$size), ]
  
  # define bin edges
  bin_edges <- c(0, 4, 5.6, 8, 11, 16, 22.6, 32, 45, 64, 90, 128, 180, 300, Inf) #when counting, includes upper limit (0-4, 4.1-5.6, etc)
  bin_labels <- c("4", "5.6", "8", "11", "16", "22.6", "32", "45", "64", "90", "128", "180", "300", "400")
  
  # assign each size value to a bin if it wasn't already binned
  df$bin <- cut(df$size, breaks = bin_edges, labels = bin_labels, right = TRUE)
  
  # count frequencies by bin
  binned_df <- aggregate(Freq ~ bin, data = df, sum, na.rm = TRUE)
  
  # Add zeroes to bins that are missing rocks
  binned_df <- data.frame(bin = bin_labels) %>%
    left_join(binned_df, by = "bin") %>%
    mutate(Freq = replace_na(Freq, 0))
  
  # calculate cumulative frequency
  binned_df$c_freq <- cumsum(binned_df$Freq)
  
  # calculate percent finer
  total_count <- sum(binned_df$Freq)
  binned_df$p_finer <- (binned_df$c_freq / total_count)
  
  # add a column for the labels for the x-axis
  bin_names <- c("<4", "<5.6", "<8", "<11", "<16", "<22.6", "<32", "<45", "<64", "<90", "<128", "<180", "<300", ">300")
  
  binned_df$bin_names <- bin_names
  
  binned_df$site <- name
  
  # rename the first column to 'size'
  colnames(binned_df)[1] <- "size"
  
  # create the output path for the new CSV file
  output_file <- paste0("pc_cleaned/", name, "_frq.csv")
  
  # write the dataframe to the new CSV file
  write.csv(binned_df, output_file, row.names = FALSE)
  
  # print which .csv is done as it finishes
  cat("Processed", name, "\n")
  
}
```
## Pre-Binned Data:
Some data was recorded as counts in the bin already, which is in a different folder.

### Make a list of the files in the binned data folder
```{r}
binned_file_list <- list.files("pc_binned", pattern = "*.csv", full.names=TRUE)
```


### Loop through each file and do the following:
```{r}
# loop through each file
for (file in binned_file_list) {
  # get the name of the file
  name <- tools::file_path_sans_ext(basename(file))
  
  # read the csv
  data <- read.csv(file, header = TRUE)
  
  # Calculate cumulative frequency
  data$c_freq <- cumsum(data$Freq)
  
  # Calculate percent finer
  total_count <- sum(data$Freq)
  data$p_finer <- (data$c_freq / total_count)
  
  # add a column for the labels for plotting
  bin_names <- c("<4", "<5.6", "<8", "<11", "<16", "<22.6", "<32", "<45", "<64", "<90", "<128", "<180", "<300", ">300")
  
  data$bin_names <- bin_names
  
  # add a column for the sites for combining with other files
  data$site <- name
  
  # create the output path for the new CSV file
  output_file <- paste0("pc_cleaned/", name, "_frq.csv")
  
  # write the dataframe to the new CSV file
  write.csv(data, output_file, row.names = FALSE)
  
  # print which .csv is done as it finishes
  cat("Processed", name, "\n")
}
```